---
title: "Reproducible Research Project 1"
author: "Allan de Venecia"
date: "May 14, 2017"
output: md_document
---

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

- Dataset: [Activity monitoring data][1]  [52K]  

The variables included in this dataset are:  

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
- date: The date on which the measurement was taken in YYYY-MM-DD format  
- interval: Identifier for the 5-minute interval in which measurement was taken  

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.  

### Loading and Preprocessing the Data

```{r results="hide"}
library(dplyr)
library(ggplot2)
# unzip(zipfile = "actiity.zip")
data <- read.csv("activity.csv")
# summary(data)
cleanData <- filter(data, !is.na(data$steps))
```

### What is mean total number of steps taken per day?

1. Remove missing dates with no steps taken from the data set. Compute for total number of steps taken per day.

```{r}
totalspd <- summarise(group_by(cleanData, date), Tsteps = sum(steps))
```

2. Create the historgram for the total number of steps taken per day
```{r}
hist(totalspd$Tsteps, breaks = 5, col = "blue", xlab = "Steps", main = "Total Steps Daily")
```

3. Calculate the mean and the median for the total steps taken daily.
```{r results="hide"}
meantspd <- as.integer(mean(totalspd$Tsteps))  
mediantspd <- median(totalspd$Tsteps)
```

Average number of steps taken daily : `r meantspd`  
Median number of steps taken daily : `r mediantspd`  

## What is the average dadily activity pattern?  

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).    

```{r}
spI <- summarise(group_by(cleanData, interval), steps = mean(steps))
dailyp <- ggplot(spI, aes(x=interval, y=steps))
dailyp+geom_line(color="blue")+xlab("Interval")+ylab("Average Steps")+ggtitle("Average Number of Steps per Interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxS <- as.integer(max(spI$steps))
maxInt <- spI[spI$steps==max(spI$steps),1]
```
  
Maximum number of steps for a 5 minute interval : `r maxS`  
5 minute interval with the maximum number of steps : `r maxInt`  

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
nullRows <- nrow(data[is.na(data$steps),])
```

Total number of rows with NA value : `r nullRows`  

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

* Fill NAs with the mean steps for each intervals 

```{r}
naData <- data[is.na(data$steps),]
naData$steps[is.na(naData$steps)] <- spI$steps[match(naData$interval, spI$interval)][which(is.na(naData$steps))]
mergeData <- rbind(cleanData, naData)
totalspd2 <- summarise(group_by(mergeData, date), Tsteps2 = sum(steps))
hist(totalspd2$Tsteps2, breaks = 5, col = "red", xlab="Steps", main="Total Steps per Day [Fixed NAs]")
hist(totalspd$Tsteps, breaks = 5, col = "green",  add=T)
legend("topright", c("Inserted Data", "Non-NA Data"), fill=c("red", "Green") )

newMean <- as.integer(mean(totalspd2$Tsteps2))
varMean <- newMean - meantspd
newMedian <- as.integer(median(totalspd2$Tsteps2))
varMed <- newMedian - mediantspd

```
 
New mean of inserted data : `r newMean`; vs original mean `r meantspd` resulting to a variance of `r varMean` average steps daly.
  
New median of inserted data : `r newMedian`; vs original mean `r mediantspd` resulting to a variance of `r varMed` steps for the median.

No real observable change for both histogram, median and mean.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
mergeData$day <- weekdays(as.Date(mergeData$date))
mergeData$dayCat <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
spI2 <- summarise(group_by(mergeData, interval, dayCat), steps = mean(steps))
dailyp2 <- ggplot(spI2, aes(x=interval, y=steps))
dailyp2+geom_line(color="blue")+facet_grid(dayCat~., scales = "free")+xlab("Interval")+ylab("Average Steps")+ggtitle("Average Steps per Interval Based on Type of Day")
```

There is an observable trend based on what day within the week the activity is done. Average steps taken seems to be higher over the weekend.
    
[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip